---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  error = FALSE,
  message = FALSE,
  warning = FALSE
)
```



# glmpa

<!-- badges: start -->
<!-- badges: end -->

The goal of glmpa is to approximate power of generalized linear mixed models using the probability distribution method (PDM).  

This goal can be achieved using this package, as it includes two functions that do this. 

These two functions both extract all the components needed to implement the PDM for a power approximation of a given fixed effect.  
The needed components include a numerator and denominator degrees of freedom, a non-centrality parameter, and an F-statistic for the given fixed effect.  


One function uses the blme (Chung et al., 2013) package to set sharp priors to fix the random variance components in order to implement the PDM.  

The other function uses the glmmTMB (Brooks et al., 2017) package through its start and map arguments to fix random variance components in order to implement the PDM.  


Implementation issues to consider include the formatting of a binomial exemplary dataset. For binomial datasets that are organized at the experimental level instead of at the subject unit level, the data must be modified to be at the subject unit level in order to extract denominator degrees of freedom. An example is included using the tidyr package (Wickham et al., 2026). 
The other implementation issue is that the fixed variable defined for power approximation has to be set as a factor in the exemplary dataset. 


The functions using the blme package builds on earlier work to approximate power of linear mixed models using the probability distribution method (Geisler 2022). 


## Installation

You can install the development version of glmpa like so:

```{r}
library("devtools") 
install_github("carlieprinster/R-Power-Approximation")  
library(glmpa)
library(glmmTMB)
library(emmeans)
```

## Example 

Using these functions first involves creating a data frame of the exemplary dataset for the model. This dataset is used  to exemplify the magnitude of the fixed effect that we are interesting in detecting.

The example is to do with an experiment that seeks to assess the effects of two treatment factors, nitrogen source and years of thatch accumulation, on chlorophyll content in grass clippings of golf greens (Kuehl 1994, example 14.1). This split plot designed experiment has an estimated response distribution from the Gaussian family. 

```{r example}
#Build Dataset
Thatch <- as.factor(c(rep(2, 16), rep(5, 16), rep(8, 16)))
NSource <- rep(c(rep("AmmSulph", 4), rep("IBDU", 4), rep("SCUrea", 4),
                 rep("Urea", 4)), 3)
estY <- c(rep(6, 12), rep(4, 4), rep(c(rep(6, 4), rep(7, 4), rep(8, 4),
                                       rep(5, 4)), 2))
Field <- as.factor(rep(1:4, 12))
ex_gaussian <- cbind.data.frame(Thatch, NSource, estY, Field)

```

# Example for blme function


One can implement the function by defining the following as arguments: model formula, the values of the random variance components, the previously created data frame, the model distribution family, the names of the fixed effects of power approximation interest, and the names of the random variance components. All other arguments have default values that can be customized as users need. 

```{r example blme}

gaussian_example <-  power_blme(Formula = estY ~ NSource*Thatch + (1 | Field) + (1|Field:NSource), 
                                  Varcomp = c(0.008, 0.07), Resid_var = 0.2, Data = ex_gaussian,
                                  Family = "gaussian", Random_Effects = c("Field", "Field:NSource"), Fixed_Effects =c("NSource", "NSource:Thatch", "Thatch"), Alpha = 0.05)

```


# Example for glmmTMB function


One can implement the other function by first defining a model within the glmmTMB package framework and defining the start and map arguments with the fixed random variance components. Then, the user must define the following: a glmmTMB model, the name of the fixed effect of power approximation interest, and the model dataset. The last argument, the alpha value, has a default value of 0.05 that can be customized as users need.   
```{r example glmmTMB}

#build model
start = list(theta = log(c(sqrt(0.008), sqrt(0.07))), betadisp = log(sqrt(0.2)))
map = list(theta = factor(c(NA, NA)), betadisp = factor(NA))

ex_gaussianmodel <- glmmTMB(estY ~ NSource * Thatch + (1 | Field) + (1 | Field:NSource),
                            data = ex_gaussian,
                            family = gaussian(),
                             control = glmmTMBControl(optimizer = optim, optArgs = list(method = "BFGS")),
                            start = start,
                            map = map)

#call function
gaussian_glmmTMB_example <- power_glmmTMB(ex_gaussianmodel, "NSource:Thatch", 0.05, ex_gaussian)

```


